Future challenges for Research Software Engineering

### 2025-05-14 future_challenges_for_RSEng.tex
minor fixes

Research Software Engineering is currently an emerging field,
with an increasing integration into various aspects of research.
In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
We present here the results of a workshop discussion at the
deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
claiming to represent the entire deRSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in imagining the future.
In the following sections, we consider the evolution of the RSE definition,
the development of RSE education,
the impact of automation and generative artificial intelligence in RSE work,
the emerging ethical considerations,
and the transformations expected in the research organizations,
and if there will be other in academia covered by new RTPs (\emph{Research Technical Professional}s).
We conclude the paper with a summary and with our wishes and recommendations.

\section{Aspects of this future}
%\subsection{What do we want to have achieved by 2035}
Looking ahead over a 10-year timeframe, we hope that the RSE community will have
contributed to make research software and its outputs
more sustainable, robust and maintainable. We see this happening in a number of ways. However,
we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.
With development of software tools and scripted workflows becoming an increasingly important part of everyday
research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
do see the work of RSEs developing and shifting as the research community changes.

@@

\subsection{Definition and Evolution of RSEs}
\begin{whatis}{What do we want to have achieved}{definition-evolution-rse}
The term RSE is established in the minds of university boards, as a valuable member in the academic framework.
RSEs are developed using specialized master's program, and their support is valued by researchers. We still teach researchers basic software skills, but the content of this lecture has changed drastically.
\end{whatis}
% Here we consider how the topics of RSEng will change in the face of the digitalization of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.
In this discussions the workshop participants found that
the digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks
(JN: for example, versioning and archiving?)
(GC: Higher-level and more intuitive workflows that require less training).
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

\begin{story}{The interaction of RSEs with AI}{RSEs_with_AI}
Kim develops advanced scientific software in tandem with an AI agent that codes, debugs, and simulates experiments alongside her. The AI proactively suggests optimizations, generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs. With this partnership, Kim accelerates her research output while maintaining full control over scientific direction and ethical oversight.
\end{story}

@@

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{What do we want to have achieved}{academic}
We expect research software to be a central part of scientific practice on the same level as data and publications.
This will be supported by the first generations of RSE-Master graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, that are known by 2035 in university boards, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
At universities which have them, we want to see that researchers naturally accept them as part of their used services.
Additionally researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks. This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years such as
the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
But we could imagine new RSE specialisations to emerge in the fields
of quantum computing, such as the Quantum-RSE, and in federated online platforms for research.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

On the other hand the existence of people with a certain specilised skillset opens up possiblities for pooling.
Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
Hopefully, these RSE-units are endowed with enough trust by the university boards that translates into stable funding.
With longer-term funding and more trust to pick up longer term projects,
some RSEs can focus on code modernization, cleanup, and modularization,
which can pay off in faster and larger development in the future.
This is an opportunity that would not be possible by short-term funded solo RSEs.

\emph{Which challenges will still persist}\\
The workshop participants expect that there will still be budgeting and administrative hurdles to creating
permanent and well-funded RSE teams.
Here it can only be strongly hoped, that, with the recognized importance of research software, state-actors
support the development with a reliable stream of funding.
This would also require a change of culture in research about the acceptance of research software.\\
While we hope that the barrier to researchers diminishes, the workshop participants
still expect some work to do. We hope that the compartmentalization that occurs will help researchers reprioritize their workload.
An upcoming challenge will be that, with the emergence of more dedicated RSE units,
networking between them will come into the foreground, because there should be a strong need to share best practices
among institutions with successful RSE Teams.
On a more organisational level, we expect the sharing of models on how how to organise diiferent RSEs, to fit the different organisational constraints.
After that Organisations will have to come up with models for hiring people.
While a standardized RSE Master's makes it easier, it will be interesting to see if
RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE master's degree ensures a hiring process on a more standardised level.

@@

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{What do we want to have achieved}{career-paths}
We would like to see the emergence of more dedicated personel, i.e. other RTPs next to the RSE.
Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
or organisations that do knowledge transfer).
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units.\cite{Kempf2025-draft}
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding is more difficult to acquire in lower levels of an organization.
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, Göttingen).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups themselves prove themselves valuable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.

\begin{story}{New career paths}{New career paths}
Ken, though trained as a physicist, found his passion not in theory, but in keeping the quantum computers humming. He became the go-to expert for troubleshooting entangled qubits and stabilizing delicate cryogenic systems. His unique blend of physics knowledge and hands-on problem solving earned him a new title among his peers: the first Quantum RTP. In a world where theory meets reality, Ken quietly ensures the future of quantum computing stays on track.
\end{story}
\section{What is in this future for RSEs and where to go from here?}
%FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
%GC-NOTE: Shall we add another story box here? Ideas what we write here?
It is hard to conclude a piece written about the future, but we have tried to list some of the trends that we foresee.
Nevertheless there are some things that we declared out of scope here.
Among that is the digital transformation in the wider society.
Digital competencies in the general population continue to increase, making jumping to software development easier than before.
But this digitalisation will also lead to changes in the way societies interact and therefore will translate into changes in how academia perceives itself.
We have also not delved into the interaction with industry much.
There are also some trends that believe to stay with us.
The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
Societies will be facing new questions, that arise from technologies such as generally available AI systems.

>>>>

Research Software Engineering is currently an emerging field,
with an increasing integration into various aspects of research.
In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
We present here the results of a workshop discussion at the
deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
claiming to represent the entire deRSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living.
In the following sections, we consider the evolution of the RSE definition,
the development of RSE education,
the growth of the RSE communities,
the impact of automation and generative artificial intelligence in RSE work,
the emerging ethical considerations,
and the transformations expected in the research organisations,
and if there will be other in academia covered by new RTPs (\emph{Research Technical Professional}s).
We conclude the paper with a summary and with our wishes and recommendations.

\section{Aspects of this future}
%\subsection{What do we want to have achieved by 2035}
Looking ahead over a 10-year timeframe, we hope that the RSE community will have
contributed to make research software and its outputs
more sustainable, robust and maintainable.
We also hope, that Research Software becomes an accepted research artifact on equal footing as data and publications.
We see this happening in a number of ways.
However, we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.
With development of software tools and scripted workflows becoming an increasingly important part of everyday
research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
do see the work of RSEs developing and shifting as the research community changes.

@@

\subsection{Definition and Evolution of RSEs}
\begin{whatis}{What do we want to have achieved}{definition-evolution-rse}
The term RSE is established in the minds of university boards, as a valuable member in the academic framework.
RSEs are developed using specialized master's program, and their support is valued by researchers. We still teach researchers basic software skills, but the content of this lecture has changed drastically.
\end{whatis}
% Here we consider how the topics of RSEng will change in the face of the digitalisation of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.
In this discussions the workshop participants found that
the digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks
(JN: for example, versioning and archiving?)
(GC: Higher-level and more intuitive workflows that require less training).
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

\begin{story}{The interaction of RSEs with AI}{RSEs_with_AI}
Kim develops advanced scientific software in tandem with an AI agent that codes, debugs, and simulates experiments alongside her. The AI proactively suggests optimisations, generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs. With this partnership, Kim accelerates her research output while maintaining full control over scientific direction and ethical oversight.
\end{story}

@@

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{What do we want to have achieved}{academic}
We expect research software to be a central part of scientific practice on the same level as data and publications.
This will be supported by the first generations of RSE-Master graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, that are known by 2035 in university boards, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
At universities which have them, we want to see that researchers naturally accept them as part of their used services.
Additionally researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks. This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years such as
the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
But we could imagine new RSE specialisations to emerge in the fields
of quantum computing, such as the Quantum-RSE, and in federated online platforms for research.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

On the other hand the existence of people with a certain specilised skillset opens up possiblities for pooling.
Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
Hopefully, these RSE-units are endowed with enough trust by the university boards that translates into stable funding.
With longer-term funding and more trust to pick up longer term projects,
some RSEs can focus on code modernisation, cleanup, and modularisation,
which can pay off in faster and larger development in the future.
This is an opportunity that would not be possible by short-term funded solo RSEs.

\emph{Which challenges will still persist}\\
The workshop participants expect that there will still be budgeting and administrative hurdles to creating
permanent and well-funded RSE teams.
Here it can only be strongly hoped, that, with the recognized importance of research software, state-actors
support the development with a reliable stream of funding.
This would also require a change of culture in research about the acceptance of research software.\\
While we hope that the barrier to researchers diminishes, the workshop participants
still expect some work to do. We hope that the compartmentalisation that occurs will help researchers reprioritize their workload.
An upcoming challenge will be that, with the emergence of more dedicated RSE units,
networking between them will come into the foreground, because there should be a strong need to share best practices
among institutions with successful RSE Teams.
On a more organisational level, we expect the sharing of models on how how to organise diiferent RSEs, to fit the different organisational constraints.
After that Organisations will have to come up with models for hiring people.
While a standardized RSE Master's makes it easier, it will be interesting to see if
RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE master's degree ensures a hiring process on a more standardised level.

@@

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{What do we want to have achieved}{career-paths}
We would like to see the emergence of more dedicated personel, i.e. other RTPs next to the RSE.
Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
or organisations that do knowledge transfer).
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units.\cite{Kempf2025-draft}
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding is more difficult to acquire in lower levels of an organisation.
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, Göttingen).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups themselves prove themselves valuable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.

\begin{story}{New career paths}{New career paths}
Ken, though trained as a physicist, found his passion not in theory, but in keeping the quantum computers humming. He became the go-to expert for troubleshooting entangled qubits and stabilizing delicate cryogenic systems. His unique blend of physics knowledge and hands-on problem solving earned him a new title among his peers: the first Quantum RTP. In a world where theory meets reality, Ken quietly ensures the future of quantum computing stays on track.
\end{story}
\section{What is in this future for RSEs and where to go from here?}
%FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
%GC-NOTE: Shall we add another story box here? Ideas what we write here?
It is hard to conclude a piece written about the future, but we have tried to list some of the trends that we foresee.
The material presented here, is based on a workshop, bt extended with the author's ideas.
Nevertheless there are some things that we declared out of scope here.
Among that is the digital transformation in the wider society.
Digital competencies in the general population continue to increase, making jumping to software development easier than before.
But this digitalisation will also lead to changes in the way societies interact and therefore will translate into changes in how academia perceives itself.
We have also not delved into the interaction with industry much.
There are also some trends that believe to stay with us.
The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
Societies will be facing new questions, that arise from technologies such as generally available AI systems.
